---
title: "Case_study"
author: "Cristian Robayna"
date: "2022-12-18"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

This analysis case study is about Capstone project requirement for Google Data Analytics Professional Certificate. For this project this steps will be followed to ensure its completion:

-   It will follow the steps of the data analysis process: Ask, prepare, process, analyze, share and act.

A brief explanation of APPAA: **Ask**

-   Ask effective questions
-   Define the scope of the analysis
-   Define what success looks like

**Prepare**

-   Verify data's integrity
-   Check data credibility and reliability
-   Check data types
-   Merge datasets

**Process**

-   Clean, Resume and Transform data
-   Document cleaning processes and results

**Analyze**

-   Identify patterns
-   Draw conclusions
-   Make preditions

**Share**

-   Create effective visuals
-   Create a story for data
-   Share insights to stakeholders

**Act**

-   Give recommendations based on insights
-   Solve problems
-   Create something new

## 1.Ask

#### Scenario

Marketing team needs to design marketing strategies aimed at converting casual riders into annual members. In order to do that, however, the marketing analyst team needs to better understand how annual members and casual riders differ.

#### Stakeholders:

-   Director of marketing
-   Cyclistic executive team

#### Objective

Hence, the objective for this analysis is to throw some light on how the two types of customers: annual members and casual riders, use Cyclistic bikeshare differently, based on few parameters that can be calculated/ obtained from existing data.

#### Deliverables:

-   Insights on how annual members and casual riders use Cyclistic bikes differently
-   Provide effective visuals and relevant data to support insights
-   Use insights to give three recommendations to convert casual riders to member riders

## 2. Prepare

#### Data Sources

12 datasets have been made available for each month starting from **January 2022 to December 2022**. Each dataset captures the details of every ride logged by the customers of Cyclistic. This data that has been made publicly available has been scrubbed to omit rider's personal information.

#### Documentation, Cleaning and Preparation of data for analysis

The combination size of all 12 datasets is nearly 1 GB. So data cleaning in spreadsheets will be inefficient compared to SQL or a programming language like R or Python. I'm choosing R simply because I could do all the process there. From preparing the data, the analysis and visualization in the same platform. As it is also an opportunity to learn better R.

#### Load libraries

```{r}
library(tidyverse)
library(ggplot2)
library(lubridate)
library(dplyr)
library(readr)
library(janitor)
library(data.table)
library(tidyr)
library(rmarkdown)
```

#### Load datasets

```{r}
tripdata_202112 <- read.csv("202112-divvy-tripdata.csv")
tripdata_202201 <- read.csv("202201-divvy-tripdata.csv")
tripdata_202202 <- read.csv("202202-divvy-tripdata.csv")
tripdata_202203 <- read.csv("202203-divvy-tripdata.csv")
tripdata_202204 <- read.csv("202204-divvy-tripdata.csv")
tripdata_202205 <- read.csv("202205-divvy-tripdata.csv")
tripdata_202206 <- read.csv("202206-divvy-tripdata.csv")
tripdata_202207 <- read.csv("202207-divvy-tripdata.csv")
tripdata_202208 <- read.csv("202208-divvy-tripdata.csv")
tripdata_202209 <- read.csv("202209-divvy-tripdata.csv")
tripdata_202210 <- read.csv("202210-divvy-tripdata.csv")
tripdata_202211 <- read.csv("202211-divvy-tripdata.csv")
```

#### Check column names of each dataset for consistency

```{r}
colnames(tripdata_202112)
colnames(tripdata_202201)
colnames(tripdata_202202)
colnames(tripdata_202203)
colnames(tripdata_202204)
colnames(tripdata_202205)
colnames(tripdata_202206)
colnames(tripdata_202207)
colnames(tripdata_202208)
colnames(tripdata_202209)
colnames(tripdata_202210)
colnames(tripdata_202211)
```

#### Check data structures and data types for all data frames

```{r}
str(tripdata_202112)
str(tripdata_202201)
str(tripdata_202202)
str(tripdata_202203)
str(tripdata_202204)
str(tripdata_202205)
str(tripdata_202206)
str(tripdata_202207)
str(tripdata_202208)
str(tripdata_202209)
str(tripdata_202210)
str(tripdata_202211)
```

Al the data seems consistent and have integrity but some of the columns don't have data as tripdata_202206, tripdata_202208 and tripdata_202209 in the columns named start_station_id, end_station_id, start_station_name and end_station_name. By checking the dataset, it is supposed that no all the cyclist start at a station, so it is blank in those cases.

#### Combine all the datasets into a single one

```{r}
all_trips <- bind_rows(tripdata_202112, tripdata_202201, tripdata_202202, tripdata_202203, tripdata_202204, tripdata_202205, tripdata_202206, tripdata_202207, tripdata_202208, tripdata_202209, tripdata_202210, tripdata_202211)
str(all_trips)
```

## 3. Process

#### Convert datatypes

started_at and ended_at should be in datetime instead of char. So we convert all the char to date time

```{r}
all_trips[['started_at']] <- ymd_hms(all_trips[['started_at']])
all_trips[['ended_at']] <- ymd_hms(all_trips[['ended_at']])

str(all_trips)
```

#### Check if there is duplicates

```{r}
nrow(all_trips[duplicated(all_trips), ])
```

#### Remove columns not required or that will not be in the scope of this project

```{r}
all_trips <- all_trips %>%
    select(-c(start_lat:end_lng))
glimpse(all_trips)
```

#### Rename columns for better readability

```{r}
all_trips <- all_trips %>%
    rename(ride_type = rideable_type, 
           start_time = started_at,
           end_time = ended_at,
           customer_type = member_casual)
glimpse(all_trips)
```

#### Add new columns for aggregate functions

```{r}
#column for day of the week the trip started
all_trips$day_of_the_week <- format(as.Date(all_trips$start_time),'%a')

#column for month when the trip started
all_trips$month <- format(as.Date(all_trips$start_time),'%b_%y')

#column for time of the day when the trip started
#Time element needs to be extracted from start_time. However, as the times must be in POSIXct
#(only times of class POSIXct are supported in ggplot2), a two-step conversion is needed. 
#First the time is converted to a character vector, effectively stripping all the date information. 
#The time is then converted back to POSIXct with today’s date – the date is of no interest to us,
#only the hours-minutes-seconds are.
all_trips$time <- format(all_trips$start_time, format = "%H:%M")
all_trips$time <- as.POSIXct(all_trips$time, format = "%H:%M")

#column for trip duration in min
all_trips$trip_duration <- (as.double(difftime(all_trips$end_time, all_trips$start_time)))/60

# check the dataframe
glimpse(all_trips)
```

Let's check to see if the trip_duration column has any negative values, as this may cause problem while creating visualizations. Also, we do not want to include the trips that were part of quality tests by the company. These trips are usually identified by string 'test' in the start_station_name column.

```{r}
# checking for trip lengths less than 0
nrow(subset(all_trips,trip_duration < 0))
  
#checking for testrides that were made by company for quality checks
nrow(subset(all_trips, start_station_name %like% "TEST"))
nrow(subset(all_trips, start_station_name %like% "test"))
nrow(subset(all_trips, start_station_name %like% "Test"))

```

As there are 100 rows with trip_dration less than 0 mins and 1 trip that was test rides, we will remove these observations from our dataframe as they contribute less than 0.01% of the total rows. We will create a new dataframe deviod of these obseravtions without making any changes to the existing dataframe.

```{r}
# remove negative trip durations 
all_trips_v2 <- all_trips[!(all_trips$trip_duration < 0),]

#remove test rides
all_trips_v2<- all_trips_v2[!((all_trips_v2$start_station_name %like% "TEST" | all_trips_v2$start_station_name %like% "test")),]


#check dataframe
glimpse(all_trips_v2)
```

It is important to make sure that customer_type column has only two distinct values. Let's confirm the same.

```{r}
# checking count of distinct values
table(all_trips_v2$customer_type)
#aggregating total trip duration by customer type
setNames(aggregate(trip_duration ~ customer_type, all_trips_v2, sum), c("customer_type", "total_trip_duration(mins)"))
```

## 4&5. Analyze and Share the Data

Our dataframe is now ready for analysis that can help us uncover some insights on how the casual and members riders use Cyclistic rideshare differently.

Let's start trying to get some simple statistics on trip_duration for all customers, and do the same by customer_type.

```{r}
# statictical summary of trip_duration for all trips
summary(all_trips_v2$trip_duration)
#statistical summary of trip_duration by customer_type
all_trips_v2 %>%
    group_by(customer_type) %>%
    summarise(min_trip_duration = min(trip_duration),max_trip_duration = max(trip_duration),
              median_trip_duration = median(trip_duration), mean_trip_duration = mean(trip_duration))
```

The mean trip duration of member riders is lower than the mean trip duration of all trips, while it is the opposite for casual riders, whose mean trip duration is higher than the mean trip duration of all trips. So, casual riders usually take the bike out for longer duration compared to members.

#### Total number of trips by costumer type and day of the week

```{r}
# fix the order for the day_of_the_week and month variable so that they show up 
# in the same sequence in output tables and visualizations
all_trips_v2$day_of_the_week <- ordered(all_trips_v2$day_of_the_week, levels=c("lu.", "ma.", "mi.", "ju.", "vi.", "sá.", "do."))
all_trips_v2$month <- ordered(all_trips_v2$month, levels=c("dic._21", "ene._22", "feb._22", "mar._22", "abr._22", "may._22", "jun._22","jul._22", "ago._22", "sep._22", "oct._22", "nov._22"))
all_trips_v2 %>% 
  group_by(customer_type, day_of_the_week) %>%  
  summarise(number_of_rides = n(),average_duration_mins = mean(trip_duration)) %>% 
  arrange(customer_type, desc(number_of_rides))
```

#### Visualizations

```{r}
all_trips_v2 %>%  
  group_by(customer_type, day_of_the_week) %>% 
  summarise(number_of_rides = n()) %>% 
  arrange(customer_type, day_of_the_week)  %>% 
  ggplot(aes(x = day_of_the_week, y = number_of_rides, fill = customer_type)) +
  labs(title ="Total trips by customer type Vs. Day of the week") +
  geom_col(width=0.5, position = position_dodge(width=0.5)) +
  scale_y_continuous(labels = function(x) format(x, scientific = FALSE))
```

From the table and graph above, casual customers are most busy on Sundays followed by Saturdays, while members are most busy on later half of the week extending into the weekend. Interesting pattern to note though is the consistent trip numbers among members with less spread over entire week as compared to casual riders who don't seem to use the bikeshare services much during weekdays.

#### Average number of trips by costumer type and month

```{r}
unique(all_trips$month)

all_trips_v2 %>% 
  group_by(customer_type, month) %>%  
  summarise(number_of_rides = n(),`average_duration_(mins)` = mean(trip_duration)) %>% 
  arrange(customer_type,desc(number_of_rides))
```

#### Visualization

```{r}
all_trips_v2 %>%  
  group_by(customer_type, month) %>% 
  summarise(number_of_rides = n()) %>% 
  arrange(customer_type, month)  %>% 
  ggplot(aes(x = month, y = number_of_rides, fill = customer_type)) +
  labs(title ="Total trips by customer type Vs. Month") +
  theme(axis.text.x = element_text(angle = 30)) +
  geom_col(width=0.5, position = position_dodge(width=0.5)) +
  scale_y_continuous(labels = function(x) format(x, scientific = FALSE))
```

The data shows that the months of July, August and September are the most busy time of the year among both members and casual riders. This could be attributed to an external factor (eg. cold weather, major quality issue) that might have hindered with customer needs. 2021 is a tough year when Covid comes. People care more about their health. The charts shows that the no.of rides in 2021 is higher than 2020 in general. However, the number of trips made by members is always higher than the casual riders across all months of the year.

#### Visualization of average trip duration by costumer type on each day of the week

```{r}
all_trips_v2 %>%  
  group_by(customer_type, day_of_the_week) %>% 
  summarise(average_trip_duration = mean(trip_duration)) %>%
  ggplot(aes(x = day_of_the_week, y = average_trip_duration, fill = customer_type)) +
  geom_col(width=0.5, position = position_dodge(width=0.5)) + 
  labs(title ="Average trip duration by customer type Vs. Day of the week")
```

The average trip duration of a casual rider is more than twice that of a member. Note that this necessarily does not mean that casual riders travel farther distance. It is also interesting to note that weekends not only contribute to more number of trips but also longer trips on average when compared to weekdays.

#### Visualization of average trip duration by customer type vs. month

```{r}
all_trips_v2 %>%  
  group_by(customer_type, month) %>% 
  summarise(average_trip_duration = mean(trip_duration)) %>%
  ggplot(aes(x = month, y = average_trip_duration, fill = customer_type)) +
  geom_col(width=0.5, position = position_dodge(width=0.5)) + 
  labs(title ="Average trip duration by customer type Vs. Month") +
  theme(axis.text.x = element_text(angle = 30))
```

Average trip duration of member riders is anywhere between 10-15 minutes throughout the year. However, there seems to be a distinct pattern when it comes to casual riders, whose average trip duration swings wildly from as low as \~20 minutes to more than 30 min depending on time of the year. It is worth noting unusually long trip durations by casual riders in the month of March.

#### Visualization of bike demand over 24hr period (a day)

```{r}
all_trips_v2 %>%  
  group_by(customer_type, time) %>% 
  summarise(number_of_trips = n()) %>%
  ggplot(aes(x = time, y = number_of_trips, color = customer_type, group = customer_type)) +
  geom_line() +
  scale_x_datetime(date_breaks = "1 hour", minor_breaks = NULL,
                   date_labels = "%H:%M", expand = c(0,0)) +
  theme(axis.text.x = element_text(angle = 90)) +
  labs(title ="Demand over 24 hours of a day", x = "Time of the day")
```

For the members, there seems to be two distict peak demand hours: 7-9 AM and 5-7 PM, the latter one coinciding with the peak demand hours of casual riders as well. One could probably hypothesize that office-goers make up majority of the members profile due to demand in both morning and evening hours, but we need more data to substantiate this assumption.

#### Visualization of ride type Vs. number of trips by costumer type

```{r}
all_trips_v2 %>%
  group_by(ride_type, customer_type) %>%
  summarise(number_of_trips = n()) %>%  
  ggplot(aes(x= ride_type, y=number_of_trips, fill= customer_type))+
              geom_bar(stat='identity') +
  scale_y_continuous(labels = function(x) format(x, scientific = FALSE)) +
  labs(title ="Ride type Vs. Number of trips")
```

Classic bikes are predominantly used by members. Docked bikes are in most demand and equally used by both members as well as casual riders.Docked bike are the least in demand of all 3 and only casual members use it. Electric bikes are more favored by members and by casuals.

#### Creating a csv file of the clean data for futher analysis or visualizations in other tools like SQL, Tableau, Power BI, etc.
```{r}
clean_data <- aggregate(all_trips_v2$trip_duration ~ all_trips_v2$customer_type + all_trips_v2$day_of_the_week, FUN = mean)
write.csv(clean_data, "Clean Data.csv", row.names = F)
```


## 6. Act

### Conclusions

* Weekend usage of bike share services is higher for casual users compared to weekday usage for members.

* Over the course of a week, the average journey length of casual riders is half of the member riders.


* While members' average journey duration are remarkably consistent month to month, casual riders ride more during the first half of the year than they do during the second.

* The most preferred bike is the electric bike followed by the classic bike.

### Recommendations

* Offer discounted membership fee for renewals after the first year. It might nudge casual members to take up membership

* Promote and upgrade the electric fleet since it seem like the most used and probably used type of vehicle in the future

## Resources

* [Stack Overflow](https://stackoverflow.com/)
* [RDocumentation](https://www.rdocumentation.org/)
* [RStudio](https://community.rstudio.com/) and [Kaggle](https://www.kaggle.com/) community

#### ------End of case study------
